---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cogcache

<!-- badges: start -->
<!-- badges: end -->

The goal of cogcache is to attempt a conversion of the GDAL warper pipeline to Rust. 

So far this project has spawned Rust crates `rwarp` and `vaster`. 

rwarp is not a geospatial I/O library but a planning and kernel library that plugs into whatever I/O layer you already have.
rwarp wants to be the warp kernel that sits inside the byte-range / obstore / VirtualiZarr / xarray ecosystem — the piece that answers "given a byte-ref store and a target grid, which bytes do I need and how do I resample them?" without owning the bytes itself.

The community this is targeting already has:

- obstore — object storage I/O
- VirtualiZarr — byte-range manifests
- xarray — n-D array model
- zarrs — Zarr V3 in Rust


What they are missing is a first-class warp planning layer that speaks their language — chunk-aligned reads, byte-range requests, fill-ratio-aware subdivision.


## Installation

You can install the development version of cogcache like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(cogcache)
## basic example code
```

## Code of Conduct
  
Please note that the cogcache project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
