---
title: "Warp Mapping Prototype: Coordinate Transform Proof"
author: "Michael Sumner"
date: "2026-03-02"
format: html
---

## Goal

Independently compute the source→destination pixel mapping that GDAL's warper
uses internally, compare against `gdalwarp`, and settle the orientation/indexing
details so we can move to chunked Rust warps with confidence.

## Setup

```{r}
#| label: setup
library(gdalraster)
library(reproj)
library(vaster)
library(slippymath)
```

## Source and destination grids

A grid spec is: CRS + extent + dimension. Everything else derives.

```{r}
#| label: grid-specs

cog_url <- "/vsicurl/https://e84-earth-search-sentinel-data.s3.us-west-2.amazonaws.com/sentinel-2-c1-l2a/55/G/CM/2026/2/S2C_T55GCM_20260227T000650_L2A/B04.tif"

ds <- new(GDALRaster, cog_url)

## Source grid: extent as c(xmin, xmax, ymin, ymax)
src_ext <- ds$bbox()[c(1, 3, 2, 4)]
src_dim <- c(ds$getRasterXSize(), ds$getRasterYSize())
src_crs <- ds$getProjectionRef()
src_gt  <- ds$getGeoTransform()
```

```{r}
#| label: dest-grid

## Slippy tile extent in EPSG:3857
tile_extent_3857 <- function(z, x, y) {
  n <- 2^z
  lon <- c(x, x + 1) / n * 360 - 180
  lat <- atan(sinh(pi * (1 - 2 * c(y, y + 1) / n))) * 180 / pi
  reproj_extent(c(lon, rev(lat)), "EPSG:3857", source = "EPSG:4326")
}

## Find the z=10 tile at the centre of our source
llex <- reproj_extent(src_ext, "EPSG:4326", source = src_crs)
z <- 10L
tile <- lonlat_to_tilenum(mean(llex[1:2]), mean(llex[3:4]), z)

dest_ext <- tile_extent_3857(z, tile$x, tile$y)
dest_dim <- c(256L, 256L)
dest_crs <- srs_to_wkt("EPSG:3857")
dest_gt  <- extent_dim_to_gt(dest_ext, dest_dim)
```

## Inverse transform: dest pixels → source pixels

For each destination pixel centre, reproject to source CRS
and convert to source pixel index.

```{r}
#| label: inverse-transform

## Destination pixel centres in destination CRS
## xy_from_cell gives all centres in cell order (row-major)
dest_xy <- xy_from_cell(dest_dim, dest_ext, seq_len(prod(dest_dim)))

## Reproject to source CRS
src_xy <- reproj_xy(dest_xy, target = src_crs, source = "EPSG:3857")

## Source pixel indices (1-based, vaster convention)
src_c <- col_from_x(src_dim, src_ext, src_xy[, 1])
src_r <- row_from_y(src_dim, src_ext, src_xy[, 2])

## Valid = inside source grid
valid <- !is.na(src_c) & !is.na(src_r) &
         src_c >= 1 & src_c <= src_dim[1] &
         src_r >= 1 & src_r <= src_dim[2]

cat("Valid:", sum(valid), "/", length(valid), "\n")
```

## Read source window and sample

```{r}
#| label: apply-map

## Bounding box of source pixels we need (0-based for gdalraster)
src_xoff  <- min(src_c[valid]) - 1L
src_yoff  <- min(src_r[valid]) - 1L
src_xsize <- max(src_c[valid]) - src_xoff
src_ysize <- max(src_r[valid]) - src_yoff

src_pixels <- ds$read(band = 1,
  xoff = src_xoff, yoff = src_yoff,
  xsize = src_xsize, ysize = src_ysize,
  out_xsize = src_xsize, out_ysize = src_ysize)

## ds$read returns row-major: pixel(0,0), pixel(1,0), ... (x varies fastest)
## matrix with byrow=TRUE: row i = scanline i, col j = pixel j
src_mat <- matrix(src_pixels, nrow = src_ysize, ncol = src_xsize, byrow = TRUE)

## Sample: adjust to window-relative 1-based indices
output <- rep(NA_real_, nrow(dest_xy))
sc <- src_c[valid] - src_xoff
sr <- src_r[valid] - src_yoff
output[valid] <- src_mat[cbind(sr, sc)]

## xy_from_cell returns row-major order (cell 1..ncol = row 1, etc.)
output_mat <- matrix(output, nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)
```

## GDAL warp to same grid

```{r}
#| label: gdal-warp

tf <- tempfile(tmpdir = "/vsimem", fileext = ".vrt")
warp(ds, tf, dest_crs,
     cl_arg = c("-ts", dest_dim,
                "-te", dest_ext[c(1, 3, 2, 4)],
                "-r", "near"))

warp_ds <- new(GDALRaster, tf)
gdal_output <- read_ds(warp_ds)
warp_ds$close()

gdal_mat <- matrix(gdal_output, nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)
```

## Compare

```{r}
#| label: compare

nodata <- ds$getNoDataValue(band = 1)
if (!is.na(nodata)) {
  output_mat[output_mat == nodata] <- NA
  gdal_mat[gdal_mat == nodata] <- NA
}

both_valid <- !is.na(output_mat) & !is.na(gdal_mat)
diff <- abs(as.numeric(output_mat) - as.numeric(gdal_mat))

cat("Pixels compared:", sum(both_valid), "\n")
cat("Identical:      ", sum(diff[both_valid] == 0), "\n")
cat("Off by ≤1:      ", sum(diff[both_valid] <= 1), "\n")
cat("Max difference: ", max(diff[both_valid]), "\n")
```

## Visual comparison

```{r}
#| label: visual
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
zlim <- quantile(c(output_mat, gdal_mat), c(0.01, 0.99), na.rm = TRUE)

ximage::ximage(output_mat, dest_ext,
  main = "Our warp (reproj + NN)", col = hcl.colors(256))
ximage::ximage(gdal_mat, dest_ext,
  main = sprintf("GDAL warp (z=%d, x=%d, y=%d)", z, tile$x, tile$y),
  col = hcl.colors(256))
```

## Orientation checks

These are the details to settle before moving to Rust.

```{r}
#| label: orientation

## Check 1: does ds$read return the same order as read_ds for a small window?
a <- ds$read(band = 1, xoff = 100, yoff = 100,
             xsize = 4, ysize = 3, out_xsize = 4, out_ysize = 3)
b <- read_ds(ds, bands = 1, xoff = 100, yoff = 100,
             xsize = 4, ysize = 3, out_xsize = 4, out_ysize = 3)[[1]]
cat("ds$read vs read_ds same window:", identical(a, b), "\n")

## Check 2: single pixel at known location
## Top-left of source (pixel 0,0) should be at geo (xmin + res/2, ymax - res/2)
tl_val <- ds$read(band = 1, xoff = 0, yoff = 0,
                  xsize = 1, ysize = 1, out_xsize = 1, out_ysize = 1)
cat("Source pixel (0,0) value:", tl_val, "\n")
cat("Geo coords:", src_gt[1] + 0.5 * src_gt[2],
    src_gt[4] + 0.5 * src_gt[6], "\n")

## Check 3: xy_from_cell ordering
## Cell 1 = top-left, cell ncol = end of first row
## So the vector is row-major: all of row 1, then row 2, ...
cat("\nxy_from_cell order:\n")
cat("  Total:", nrow(dest_xy), "=", dest_dim[1], "x", dest_dim[2], "\n")
cat("  First 4 (should be first row, x increasing):\n")
print(head(dest_xy, 4))
cat("  Cell", dest_dim[1], "(end of row 1):\n")
print(dest_xy[dest_dim[1], ])
cat("  Cell", dest_dim[1] + 1, "(start of row 2, same x as cell 1, y decreased):\n")
print(dest_xy[dest_dim[1] + 1, ])

## Check 4: single dest pixel end-to-end
## Pick dest pixel at centre of the tile (cell index)
mid_cell <- (dest_dim[2] %/% 2) * dest_dim[1] + dest_dim[1] %/% 2
cat("\nDest cell", mid_cell, "(centre of tile):\n")
cat("  Dest coord:", dest_xy[mid_cell, ], "\n")
cat("  Src coord: ", src_xy[mid_cell, ], "\n")
cat("  Src pixel: ", src_c[mid_cell], src_r[mid_cell], "(1-based)\n")
cat("  Our value: ", output[mid_cell], "\n")
cat("  Direct read:", ds$read(band = 1,
    xoff = src_c[mid_cell] - 1L, yoff = src_r[mid_cell] - 1L,
    xsize = 1, ysize = 1, out_xsize = 1, out_ysize = 1), "\n")

## Check 5: what does GDAL give for that same dest pixel?
mid_row <- mid_cell %/% dest_dim[1] + 1L
mid_col <- mid_cell %% dest_dim[1]
if (mid_col == 0) { mid_col <- dest_dim[1]; mid_row <- mid_row - 1L }
cat("  GDAL value:", gdal_mat[mid_row, mid_col], "\n")
```

## Cleanup

```{r}
#| label: cleanup
ds$close()
```

## What this proves

1. **The transform map is a data structure** — for each destination pixel,
   we have explicit source pixel coordinates. This is what GDAL computes
   inside `GDALGenImgProjTransform` but never exposes.

2. **Nearest neighbour is trivial once you have the map** — the hard part
   is the coordinate transform, not the sampling.

3. **The map is independent of pixel data** — compute it once, apply to
   any band or timestep. This is the chunk→fragment mapping.

4. **GDAL agreement** — our independent coordinate transform produces
   visually identical output. Remaining pixel-level differences are from
   GDAL's ApproxTransform interpolation vs our exact per-pixel PROJ calls.

## Next: Rust warp kernel

```rust
pub fn compute_warp_map(
    src_crs: &str,        // WKT or EPSG
    src_gt: [f64; 6],     // geotransform
    src_dim: [usize; 2],  // (width, height)
    dst_crs: &str,
    dst_gt: [f64; 6],
    dst_dim: [usize; 2],
) -> WarpMap

pub struct WarpMap {
    pub dst_dim: [usize; 2],
    pub src_cols: Vec<i32>,   // -1 for out-of-bounds
    pub src_rows: Vec<i32>,   // -1 for out-of-bounds
}
```

Grid arithmetic (`extent_dim_to_gt`, `x_from_col`, `col_from_x`) becomes
equivalent Rust functions — same semantics as vaster, not transliterated GDAL C++.
