---
title: "Warp Mapping Prototype: Coordinate Transform Proof"
author: "Michael Sumner"
date: "2026-03-02"
format: html
---

## Goal

Independently compute the source→destination pixel mapping that GDAL's warper
uses internally, compare against `gdalwarp`, and establish the contracts for
moving to chunked Rust warps.

## Setup

```{r}
#| label: setup
library(gdalraster)
library(reproj)
library(vaster)
library(slippymath)
```

## Source and destination grids

A grid spec is: CRS + extent + dimension. Everything else derives.

```{r}
#| label: grid-specs

cog_url <- "/vsicurl/https://e84-earth-search-sentinel-data.s3.us-west-2.amazonaws.com/sentinel-2-c1-l2a/55/G/CM/2026/2/S2C_T55GCM_20260227T000650_L2A/B04.tif"

ds <- new(GDALRaster, cog_url)

src_ext <- ds$bbox()[c(1, 3, 2, 4)]
src_crs <- ds$getProjectionRef()
src_gt  <- ds$getGeoTransform()
```

```{r}
#| label: dest-grid

tile_extent_3857 <- function(z, x, y) {
  n <- 2^z
  lon <- c(x, x + 1) / n * 360 - 180
  lat <- atan(sinh(pi * (1 - 2 * c(y, y + 1) / n))) * 180 / pi
  reproj_extent(c(lon, rev(lat)), "EPSG:3857", source = "EPSG:4326")
}

llex <- reproj_extent(src_ext, "EPSG:4326", source = src_crs)
z <- 10L
tile <- lonlat_to_tilenum(mean(llex[1:2]), mean(llex[3:4]), z)

dest_ext <- tile_extent_3857(z, tile$x, tile$y)
dest_dim <- c(256L, 256L)
dest_crs <- srs_to_wkt("EPSG:3857")
dest_gt  <- extent_dim_to_gt(dest_ext, dest_dim)
```

## Our warp: reproject dest pixel centres → extract from source

For each destination pixel centre, `pixel_extract` reprojects to source CRS
and reads the value at that location. This is the complete warp in two lines.

```{r}
#| label: our-warp

dest_xy <- xy_from_cell(dest_dim, dest_ext, seq_len(prod(dest_dim)))
src_pixels <- pixel_extract(ds, bands = 1L, dest_xy, xy_srs = dest_crs, interp = "near")
```

## GDAL warp to same grid

```{r}
#| label: gdal-warp

tf <- tempfile(tmpdir = "/vsimem", fileext = ".vrt")
warp(ds, tf, dest_crs,
     cl_arg = c("-ts", dest_dim,
                "-te", dest_ext[c(1, 3, 2, 4)],
                "-r", "near"))

warp_ds <- new(GDALRaster, tf)
gdal_output <- read_ds(warp_ds)
warp_ds$close()
```

## Compare flat vectors (before any matrix reshaping)

```{r}
#| label: compare

nodata <- ds$getNoDataValue(band = 1)

our <- src_pixels
gdal <- gdal_output

if (!is.na(nodata)) {
  our[our == nodata] <- NA
  gdal[gdal == nodata] <- NA
}

both_valid <- !is.na(our) & !is.na(gdal)
diff <- abs(our - gdal)

cat("Pixels compared:", sum(both_valid), "\n")
cat("Identical:      ", sum(diff[both_valid] == 0), "\n")
cat("Off by ≤1:      ", sum(diff[both_valid] <= 1), "\n")
cat("Max difference: ", max(diff[both_valid]), "\n")
```

## Visual comparison

```{r}
#| label: visual
#| fig-width: 10
#| fig-height: 5

## matrix(, nrow = nrows, byrow = TRUE) for ximage display only
src_mat  <- matrix(our,  nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)
gdal_mat <- matrix(gdal, nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)

par(mfrow = c(1, 2))
zlim <- quantile(c(src_mat, gdal_mat), c(0.01, 0.99), na.rm = TRUE)

ximage::ximage(src_mat, dest_ext,
  main = "Our warp (pixel_extract + NN)", col = hcl.colors(256))
ximage::ximage(gdal_mat, dest_ext,
  main = sprintf("GDAL warp (z=%d, x=%d, y=%d)", z, tile$x, tile$y),
  col = hcl.colors(256))
```

```{r}
#| label: scatter
#| fig-width: 6
#| fig-height: 6

plot(c(src_mat), c(gdal_mat), pch = ".", cex = 0.5,
     xlab = "Our warp", ylab = "GDAL warp",
     main = "Pixel value comparison")
abline(0, 1, col = "red")
```

## Cleanup

```{r}
#| label: cleanup
ds$close()
```

## What this proves

1. **The warp is two lines** — `xy_from_cell` + `pixel_extract` gives us the
   complete nearest-neighbour warp from any dest grid to any source.

2. **Visually identical to GDAL** — same features, same orientation, same extent.

3. **Pixel-level differences remain** — not from orientation or scale/offset
   issues. The source of disagreement between `pixel_extract` (exact per-pixel
   PROJ) and `warp` (GDAL's internal pipeline) is still not fully understood.
   `-et 0` made no difference. Bilinear shows similar patterns. This is worth
   investigating further but is not blocking the Rust work.

4. **Contracts established**:
   - `xy_from_cell(dim, ext, 1:n)` returns row-major cell centres
   - `ds$read` / `read_ds` return row-major (x varies fastest)
   - `matrix(data, nrow = nrows, byrow = TRUE)` for ximage display
   - Compare flat vectors before matrix reshaping
   - gdalraster returns raw (unscaled) values; scale/offset from
     `ds$getScale()` / `ds$getOffset()` if needed

## Next: Rust warp kernel

The R warp is:

```r
dest_xy    <- xy_from_cell(dest_dim, dest_ext, 1:n)
src_xy     <- reproj_xy(dest_xy, target = src_crs, source = dest_crs)
src_pixels <- pixel_extract(ds, 1L, dest_xy, xy_srs = dest_crs, interp = "near")
```

The Rust equivalent:

```rust
pub fn compute_warp_map(
    src_crs: &str,        // WKT or EPSG
    src_gt: [f64; 6],     // geotransform
    src_dim: [usize; 2],  // (width, height)
    dst_crs: &str,
    dst_gt: [f64; 6],
    dst_dim: [usize; 2],
) -> WarpMap

pub struct WarpMap {
    pub dst_dim: [usize; 2],
    pub src_cols: Vec<i32>,   // -1 for out-of-bounds
    pub src_rows: Vec<i32>,
}
```

Grid arithmetic mirrors vaster semantics (dim, ext argument order; 1-based
in R, 0-based in Rust). Coordinate transforms via `proj` crate. Pixel
sampling from decoded tile bytes (already working in cogcache).
