---
title: "Warp Mapping Prototype: Coordinate Transform Proof"
author: "Michael Sumner"
date: "2026-03-01"
format: html
---

## Goal

Prove that we can independently compute the source→destination pixel mapping
that GDAL's warper uses internally. Given a source grid (UTM, from the COG)
and a destination grid (e.g. EPSG:3857 web mercator tile), compute which source
pixel each destination pixel maps to, then sample with nearest neighbour and
compare against `gdalwarp`.

This is the coordinate transform core — the piece nobody has extracted as a
standalone module.

## Setup

```{r}
#| label: setup
library(gdalraster)
library(grout)
library(reproj)
library(vaster)
```

## Step 1: Define source and destination grids

A "grid spec" is the minimal description of a regular raster grid:
CRS + extent + dimensions. Everything else (resolution, geotransform,
pixel centres vs corners) derives from these three.

```{r}
#| label: grid-specs

## Source: our Sentinel-2 COG in UTM zone 55S
cog_url <- "/vsicurl/https://e84-earth-search-sentinel-data.s3.us-west-2.amazonaws.com/sentinel-2-c1-l2a/55/G/CM/2026/2/S2C_T55GCM_20260227T000650_L2A/B04.tif"

ds <- new(GDALRaster, cog_url)

src_grid <- list(
  crs    = ds$getProjectionRef(),
  extent = ds$bbox()[c(1, 3, 2, 4)],          # c(xmin, xmax, ymin, ymax)
  dim    = c(ds$getRasterXSize(), ds$getRasterYSize()),
  gt     = ds$getGeoTransform()
)

cat("Source grid:\n")
cat("  CRS:   ", substr(src_grid$crs, 1, 60), "...\n")
cat("  Extent:", src_grid$extent, "\n")
cat("  Dim:   ", src_grid$dim, "\n")
cat("  Res:   ", src_grid$gt[2], src_grid$gt[6], "\n")

```

Now define a destination grid. For this proof, let's use a web mercator
tile that overlaps our source. We'll compute the extent from a z/x/y
tile address.

```{r}
#| label: dest-grid

## Helper: compute web mercator tile extent from z/x/y
## Based on OSM/slippy map tile conventions
tile_extent_3857 <- function(z, x, y) {
  n <- 2^z
  lon_min <- x / n * 360 - 180
  lon_max <- (x + 1) / n * 360 - 180
  lat_max <- atan(sinh(pi * (1 - 2 * y / n))) * 180 / pi
  lat_min <- atan(sinh(pi * (1 - 2 * (y + 1) / n))) * 180 / pi

  print(c(lon_min, lon_max, lat_min, lat_max))
  reproj::reproj_extent(c(lon_min, lon_max, lat_min, lat_max), "EPSG:3857",
                         source = "EPSG:4326")
}


llex <- reproj::reproj_extent(src_grid$extent, "EPSG:4326", source = src_grid$crs)

z <- 10L
tile <- slippymath::lonlat_to_tilenum(mean(llex[1:2]), mean(llex[3:4]), 10)
tx <- tile$x; ty <- tile$y
dest_extent <- tile_extent_3857(z, tx, ty)

## Standard web tile is 256x256
dest_dim <- c(256L, 256L)

dest_grid <- list(
  crs    = gdalraster::srs_to_wkt("EPSG:3857"),
  extent = dest_extent,
  dim    = dest_dim,
  gt     = vaster::extent_dim_to_gt(dest_extent, dest_dim)
)
names(dest_grid$gt) <- NULL

cat("Destination grid (z/x/y =", z, tx, ty, "):\n")
cat("  Extent:", dest_grid$extent, "\n")
cat("  Dim:   ", dest_grid$dim, "\n")
cat("  Res:   ", dest_grid$gt[2], dest_grid$gt[6], "\n")
```

## Step 2: Build the inverse transform map

For each destination pixel, compute its coordinate in destination CRS,
transform to source CRS, then find the source pixel. This is the inverse
transform — the heart of the warper.

```{r}
#| label: inverse-transform

## Generate destination pixel centre coordinates
## Pixel (col, row) -> geo coordinate via geotransform
dest_cols <- seq_len(dest_dim[1]) - 0.5  # pixel centres
dest_rows <- seq_len(dest_dim[2]) - 0.5

## Full grid of destination pixel centres
dest_xy <- expand.grid(col = dest_cols, row = dest_rows)
dest_x <- dest_grid$gt[1] + dest_xy$col * dest_grid$gt[2]
dest_y <- dest_grid$gt[4] + dest_xy$row * dest_grid$gt[6]

## Transform from dest CRS (EPSG:3857) to source CRS (UTM)
## reproj_xy takes a matrix and does the batch transform directly
src_coords <- reproj::reproj_xy(
  cbind(dest_x, dest_y),
  target = src_grid$crs,
  source = "EPSG:3857"
)

## Convert source coordinates to source pixel indices
## Inverse geotransform: pixel = (geo - origin) / resolution
src_col <- (src_coords[, 1] - src_grid$gt[1]) / src_grid$gt[2]
src_row <- (src_coords[, 2] - src_grid$gt[4]) / src_grid$gt[6]

## Nearest neighbour: floor to get pixel index (0-based)
src_col_nn <- floor(src_col)
src_row_nn <- floor(src_row)

## Build the transform map
transform_map <- data.frame(
  dest_col = as.integer(dest_xy$col - 0.5),  # 0-based
  dest_row = as.integer(dest_xy$row - 0.5),
  src_col  = as.integer(src_col_nn),
  src_row  = as.integer(src_row_nn),
  src_x    = src_coords[, 1],
  src_y    = src_coords[, 2]
)

## Flag out-of-bounds pixels
transform_map$valid <- transform_map$src_col >= 0 &
                       transform_map$src_col < src_grid$dim[1] &
                       transform_map$src_row >= 0 &
                       transform_map$src_row < src_grid$dim[2]

cat("Transform map:\n")
cat("  Total dest pixels:", nrow(transform_map), "\n")
cat("  Valid (in source): ", sum(transform_map$valid), "\n")
cat("  Out of bounds:     ", sum(!transform_map$valid), "\n")
cat("  Source col range:  ", range(transform_map$src_col[transform_map$valid]), "\n")
cat("  Source row range:  ", range(transform_map$src_row[transform_map$valid]), "\n")
```

## Step 3: Apply the map (nearest neighbour sampling)

Read the source pixels and apply the transform map.

```{r}
#| label: apply-map

## Which source tiles do we need? The transform map tells us.
valid_map <- transform_map[transform_map$valid, ]

## Source pixel range we need to read
src_xoff  <- min(valid_map$src_col)
src_yoff  <- min(valid_map$src_row)
src_xsize <- max(valid_map$src_col) - src_xoff + 1L
src_ysize <- max(valid_map$src_row) - src_yoff + 1L

cat("Reading source window:", src_xoff, src_yoff, src_xsize, src_ysize, "\n")

## Read the source window
src_pixels <- ds$read(
  band = 1,
  xoff = src_xoff, yoff = src_yoff,
  xsize = src_xsize, ysize = src_ysize,
  out_xsize = src_xsize, out_ysize = src_ysize
)

## Build a matrix for easy indexing (row-major)
src_mat <- matrix(src_pixels, nrow = src_ysize, ncol = src_xsize, byrow = TRUE)

## Sample: for each valid dest pixel, look up the source pixel
## Adjust indices relative to the window we read
output <- rep(NA_real_, nrow(transform_map))
for (i in which(transform_map$valid)) {
  sc <- transform_map$src_col[i] - src_xoff + 1L
  sr <- transform_map$src_row[i] - src_yoff + 1L
  output[i] <- src_mat[sr, sc]
}

## Reshape to destination grid
output_mat <- matrix(output, nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)
```

## Step 4: Compare with GDAL's warp

```{r}
#| label: gdal-warp


gdalraster::warp(ds, tf <- tempfile(tmpdir = "/vsimem", fileext = ".vrt"), dest_grid$crs, 
                 cl_arg = c("-ts", dest_grid$dim, "-te", dest_grid$extent[c(1, 3, 2, 4)]))
warp_ds <- new(GDALRaster, tf)
gdal_output <- read_ds(warp_ds)
warp_ds$close()

gdal_mat <- matrix(gdal_output, nrow = dest_dim[2], ncol = dest_dim[1], byrow = TRUE)
ximage::ximage(gdal_mat, dest_grid$extent)
```

## Step 5: Compare the results

```{r}
#| label: compare

## Apply nodata
nodata <- ds$getNoDataValue(band = 1)
if (!is.na(nodata)) {
  output_mat[output_mat == nodata] <- NA
  gdal_mat[gdal_mat == nodata] <- NA
}

## Exact match?
exact_match <- identical(as.numeric(output_mat), as.numeric(gdal_mat))
cat("Exact match:", exact_match, "\n")

## If not exact, how close?
if (!exact_match) {
  both_valid <- !is.na(output_mat) & !is.na(gdal_mat)
  diff <- abs(as.numeric(output_mat) - as.numeric(gdal_mat))
  cat("Pixels compared:", sum(both_valid), "\n")
  cat("Identical:      ", sum(diff[both_valid] == 0), "\n")
  cat("Off by 1:       ", sum(diff[both_valid] == 1), "\n")
  cat("Max difference: ", max(diff[both_valid]), "\n")

  ## Off-by-one is expected at pixel boundaries due to
  ## rounding differences (floor vs round) in the coordinate transform
}
```

## Step 6: Visual comparison

```{r}
#| label: visual-warp
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))
zlim <- quantile(c(output_mat, gdal_mat), c(0.01, 0.99), na.rm = TRUE)

ximage::ximage(output_mat,
  main = "Our warp (R transform map + NN)",
  zlim = zlim, col = hcl.colors(256), axes = FALSE
)
ximage::ximage(gdal_mat,
  main = sprintf("GDAL warp (z=%d, x=%d, y=%d)", z, tx, ty),
  zlim = zlim, col = hcl.colors(256), axes = FALSE
)
```

## Step 7: Visualise the transform map itself

```{r}
#| label: vis-transform-map
#| fig-width: 10
#| fig-height: 5

## The source column for each dest pixel — shows the warp distortion
src_col_mat <- matrix(transform_map$src_col, nrow = dest_dim[2],
                      ncol = dest_dim[1], byrow = TRUE)
src_row_mat <- matrix(transform_map$src_row, nrow = dest_dim[2],
                      ncol = dest_dim[1], byrow = TRUE)

par(mfrow = c(1, 2))
ximage::ximage(src_col_mat,
  main = "Source column index per dest pixel",
  col = hcl.colors(256, "Viridis"), axes = FALSE
)
ximage::ximage(src_row_mat,
  main = "Source row index per dest pixel",
  col = hcl.colors(256, "Viridis"), axes = FALSE
)
```

## Cleanup

```{r}
#| label: cleanup
ds$close()
```

## What this proves

1. **The transform map is a data structure** — for each destination pixel,
   we have explicit (src_col, src_row) coordinates. This is what GDAL computes
   inside `GDALGenImgProjTransform` but never exposes.

2. **Nearest neighbour is trivial once you have the map** — floor the
   coordinates. The hard part is the coordinate transform, not the sampling.

3. **The map is independent of pixel data** — you can compute it once and
   apply it to any band, any timestep. This is the chunk→fragment mapping
   that ODC builds implicitly.

4. **GDAL agreement** — our independent coordinate transform produces the
   same (or very nearly the same) warped output as `gdalwarp`.

## What's next for Rust

The Rust function signature for the core warp mapping:

```
pub fn compute_warp_map(
    src_crs: &str,        // WKT or EPSG
    src_gt: [f64; 6],     // geotransform
    src_dim: [usize; 2],  // (width, height)
    dst_crs: &str,        // WKT or EPSG
    dst_gt: [f64; 6],     // geotransform
    dst_dim: [usize; 2],  // (width, height)
) -> WarpMap {
    // For each dst pixel, compute src pixel (nearest neighbour)
    // Uses PROJ for the coordinate transform
}

pub struct WarpMap {
    pub dst_dim: [usize; 2],
    pub src_cols: Vec<i32>,   // -1 for out-of-bounds
    pub src_rows: Vec<i32>,   // -1 for out-of-bounds
}
```

This is the extracted, standalone module. Apply it to decoded tile pixels
and you have a warped output. The `proj` crate (or `proj4rs` for WASM)
provides the coordinate transforms.

## Performance notes

The R version transforms 65,536 points (256×256) through reproj/PROJ as a
batch matrix operation — already much faster than per-point sf transforms.
In Rust with the `proj` crate, this would be similar — a batch transform on
a contiguous buffer. The ApproxTransform trick from GDAL (transform a few
points, interpolate the rest) would make it even faster for smooth transforms
like UTM↔Mercator where the mapping is locally affine.
