---
title: "COG → DuckDB Cache Prototype"
author: "Michael Sumner"
date: "2026-03-01"
format: html
---

## Goal

Prove that a Cloud Optimised GeoTIFF's internal tile structure can be
faithfully catalogued in DuckDB, compressed tile bytes fetched via HTTP range
requests and stored as BLOBs, then independently decoded to produce
pixel-identical results to GDAL. This is the evidence layer for a Rust-native
warper architecture that replaces GDAL's opaque block cache with a queryable,
portable, inspectable tile store.

The deliverable is a `.duckdb` file containing the full tile catalog and raw
compressed bytes — hand it to someone and say "this *is* that COG."

## Setup

```{r}
#| label: setup
library(gdalraster)
library(grout)
library(duckdb)
library(httr2)
library(blob)
```

## Pick a COG

We need a DEFLATE-compressed, UInt16, tiled Sentinel-2 COG. Element 84's Earth
Search hosts these on S3 in two collections:

- `sentinel-2-l2a` (older): bucket `sentinel-cogs`
- `sentinel-2-c1-l2a` (newer, baseline 5.0+): bucket `e84-earth-search-sentinel-data`

Use `rstac` or the STAC browser to find a clear-sky scene for your area.

```{r}
#| label: pick-cog

## Edit this URL to your chosen COG
cog_url <- "/vsicurl/https://e84-earth-search-sentinel-data.s3.us-west-2.amazonaws.com/sentinel-2-c1-l2a/55/G/CM/2026/2/S2C_T55GCM_20260227T000650_L2A/B04.tif"

## Open and verify
ds <- new(GDALRaster, cog_url)
cat("Dimensions:", ds$getRasterXSize(), "x", ds$getRasterYSize(), "\n")
cat("Block size:", ds$getBlockSize(band = 1), "\n")
cat("Structure: ", ds$getMetadata(band = 0, domain = "IMAGE_STRUCTURE"), "\n")
cat("Data type: ", ds$getDataTypeName(band = 1), "\n")
cat("Nodata:    ", ds$getNoDataValue(band = 1), "\n")
```

Confirm COMPRESSION=DEFLATE, PREDICTOR=2, UInt16. If LZW or predictor=3,
find a different COG.

## Step 1: Build the tile catalog

grout gives us the tile grid geometry from dimension + block size + extent.
gdalraster gives us the byte offsets via the GTiff driver's `BLOCK_OFFSET`
metadata items in the `"TIFF"` domain.

```{r}
#| label: tile-catalog

dim_xy     <- c(ds$getRasterXSize(), ds$getRasterYSize())
block_size <- ds$getBlockSize(band = 1)
gt         <- ds$getGeoTransform()
crs_wkt    <- ds$getProjectionRef()
bbox       <- ds$bbox()

n_tiles_x <- ceiling(dim_xy[1] / block_size[1])
n_tiles_y <- ceiling(dim_xy[2] / block_size[2])

## grout tile grid — handles extents, edge tiles, everything
gtile  <- grout(dim_xy, bbox[c(1, 3, 2, 4)], block_size)
gindex <- tile_index(gtile)

cat(sprintf("%d x %d = %d tiles\n", n_tiles_x, n_tiles_y, n_tiles_x * n_tiles_y))
```

## Step 2: Get byte offsets from GDAL

The GTiff driver exposes per-block byte offsets as band-level metadata.
Each tile has `BLOCK_OFFSET_{x}_{y}` (file position) and
`BLOCK_SIZE_{x}_{y}` (compressed byte count).

```{r}
#| label: byte-offsets

get_block_byte_ref <- function(ds, band, tile_col, tile_row) {
  offset_key <- sprintf("BLOCK_OFFSET_%d_%d", tile_col, tile_row)
  size_key   <- sprintf("BLOCK_SIZE_%d_%d", tile_col, tile_row)

  offset <- ds$getMetadataItem(band = band, mdi_name = offset_key, domain = "TIFF")
  bcount <- ds$getMetadataItem(band = band, mdi_name = size_key, domain = "TIFF")

  list(
    byte_offset = if (nzchar(offset)) as.numeric(offset) else NA_real_,
    byte_length = if (nzchar(bcount)) as.integer(bcount) else NA_integer_
  )
}

## Quick test
get_block_byte_ref(ds, band = 1, tile_col = 0, tile_row = 0)
```

Build the full catalog with byte refs and load into DuckDB:

```{r}
#| label: build-catalog

tile_catalog <- expand.grid(
  tile_col = seq_len(n_tiles_x) - 1L,
  tile_row = seq_len(n_tiles_y) - 1L
)

## Pixel extents
tile_catalog$pixel_x_off  <- tile_catalog$tile_col * block_size[1]
tile_catalog$pixel_y_off  <- tile_catalog$tile_row * block_size[2]
tile_catalog$pixel_x_size <- pmin(block_size[1], dim_xy[1] - tile_catalog$pixel_x_off)
tile_catalog$pixel_y_size <- pmin(block_size[2], dim_xy[2] - tile_catalog$pixel_y_off)

## Geo extents from geotransform
tile_catalog$geo_xmin <- gt[1] + tile_catalog$pixel_x_off * gt[2]
tile_catalog$geo_xmax <- gt[1] + (tile_catalog$pixel_x_off + tile_catalog$pixel_x_size) * gt[2]
tile_catalog$geo_ymax <- gt[4] + tile_catalog$pixel_y_off * gt[6]
tile_catalog$geo_ymin <- gt[4] + (tile_catalog$pixel_y_off + tile_catalog$pixel_y_size) * gt[6]

## Byte offsets from GDAL's TIFF metadata domain
byte_refs <- mapply(
  function(tc, tr) get_block_byte_ref(ds, band = 1, tc, tr),
  tile_catalog$tile_col, tile_catalog$tile_row,
  SIMPLIFY = FALSE
)
tile_catalog$byte_offset <- vapply(byte_refs, `[[`, numeric(1), "byte_offset")
tile_catalog$byte_length <- vapply(byte_refs, `[[`, integer(1), "byte_length")

tile_catalog$source_uri <- cog_url
tile_catalog$ifd_index  <- 0L

## Into DuckDB
con <- dbConnect(duckdb())
dbWriteTable(con, "tile_catalog", tile_catalog)

## Sanity check
dbGetQuery(con, "
  SELECT count(*) as n_tiles,
         count(byte_offset) as n_with_offsets,
         sum(byte_length) as total_compressed_bytes,
         min(byte_offset) as first_offset,
         max(byte_offset + byte_length) as last_byte
  FROM tile_catalog
")
```

## Step 3: Fetch and store compressed tile bytes

HTTP range requests fetch raw compressed bytes directly from S3. We store them
as BLOBs in DuckDB.

```{r}
#| label: fetch-bytes

fetch_tile_bytes <- function(url, byte_offset, byte_length) {
  http_url <- sub("^/vsicurl/", "", url)

  req <- request(http_url) |>
    req_headers(
      Range = sprintf("bytes=%d-%d", byte_offset, byte_offset + byte_length - 1)
    )

  resp <- req_perform(req)
  resp_body_raw(resp)
}

## Add blob column
dbExecute(con, "ALTER TABLE tile_catalog ADD COLUMN raw_bytes BLOB")

## Fetch a random subset — enough to prove the concept
sample_tiles <- dbGetQuery(con, "
  SELECT tile_col, tile_row, byte_offset, byte_length
  FROM tile_catalog
  WHERE byte_offset IS NOT NULL
  ORDER BY random()
  LIMIT 20
")
```

```{r}
#| label: store-blobs

fetched <- lapply(seq_len(nrow(sample_tiles)), function(i) {
  raw <- fetch_tile_bytes(
    cog_url,
    sample_tiles$byte_offset[i],
    sample_tiles$byte_length[i]
  )
  list(
    tile_col  = sample_tiles$tile_col[i],
    tile_row  = sample_tiles$tile_row[i],
    raw_bytes = raw
  )
})

## Build blob data.frame — must use as_blob(), not blob()
raw_list <- lapply(fetched, function(x) x$raw_bytes)

blob_df <- data.frame(
  tile_col = vapply(fetched, `[[`, integer(1), "tile_col"),
  tile_row = vapply(fetched, `[[`, integer(1), "tile_row")
)
blob_df$raw_bytes <- as_blob(raw_list)

## Bulk write + update
dbWriteTable(con, "tile_blobs", blob_df, overwrite = TRUE)

dbExecute(con, "
  UPDATE tile_catalog SET raw_bytes = tile_blobs.raw_bytes
  FROM tile_blobs
  WHERE tile_catalog.tile_col = tile_blobs.tile_col
    AND tile_catalog.tile_row = tile_blobs.tile_row
    AND tile_catalog.ifd_index = 0
")
```

```{r}
#| label: verify-blobs

## blob_size must equal byte_length for every row
dbGetQuery(con, "
  SELECT tile_col, tile_row,
         octet_length(raw_bytes) as blob_size,
         byte_length
  FROM tile_catalog
  WHERE raw_bytes IS NOT NULL
  LIMIT 10
")
```

## Step 4: Decode and compare with GDAL

`vrt://?block=i,j` (GDAL 3.13) reads a single block via
`GDALRasterBand::ReadBlock()`. We decode our cached bytes independently and
compare.

```{r}
#| label: gdal-block-read

## Read a specific block via vrt://?block= (GDAL 3.13)
read_block_via_gdal <- function(cog_url, tile_col, tile_row) {
  block_url <- sprintf("vrt://%s?block=%d,%d", cog_url, tile_col, tile_row)
  ds_block <- new(GDALRaster, block_url)
  nx <- ds_block$getRasterXSize()
  ny <- ds_block$getRasterYSize()
  pix <- ds_block$read(band = 1, xoff = 0, yoff = 0,
                       xsize = nx, ysize = ny,
                       out_xsize = nx, out_ysize = ny)
  ds_block$close()
  pix
}
```

The decode pipeline: DEFLATE decompress → read as UInt16 → undo predictor=2
horizontal differencing via cumsum per row.

```{r}
#| label: decode-cached

## Decode DEFLATE-compressed tile bytes with predictor=2 undo
##
## Predictor=2 stores horizontal differences between consecutive samples.
## Recovery: cumsum each row, mod 2^16 for UInt16 wraparound.
## memDecompress(..., type = "gzip") handles TIFF's zlib-wrapped DEFLATE.
decode_deflate_tile <- function(raw_bytes, tile_width, tile_height,
                                nbytes = 2L, predictor = 2L) {
  decompressed <- memDecompress(raw_bytes, type = "gzip")

  pix <- readBin(decompressed, what = "integer", n = tile_width * tile_height,
                 size = nbytes, signed = FALSE, endian = "little")

  if (predictor == 1L) return(pix)

  ## Predictor == 2: sample-level horizontal differencing
  modulus <- 2L^(nbytes * 8L)
  mat <- matrix(pix, nrow = tile_height, ncol = tile_width, byrow = TRUE)
  for (i in seq_len(nrow(mat))) {
    mat[i, ] <- cumsum(mat[i, ]) %% modulus
  }
  as.integer(t(mat))
}
```

```{r}
#| label: compare-tiles

## Metadata for decode
predictor_str <- ds$getMetadataItem(band = 0, mdi_name = "PREDICTOR",
                                     domain = "IMAGE_STRUCTURE")
predictor <- if (nzchar(predictor_str)) as.integer(predictor_str) else 1L
nodata    <- ds$getNoDataValue(band = 1)
cat("Predictor:", predictor, " Nodata:", nodata, "\n")

## Compare our decode with GDAL's read for each cached tile
results <- data.frame(tile_col = integer(), tile_row = integer(),
                      match = logical())

for (i in seq_len(nrow(sample_tiles))) {
  tc <- sample_tiles$tile_col[i]
  tr <- sample_tiles$tile_row[i]

  ## Actual tile dimensions (edge tiles may be smaller than block_size)
  actual_w <- min(block_size[1], dim_xy[1] - tc * block_size[1])
  actual_h <- min(block_size[2], dim_xy[2] - tr * block_size[2])

  ## Our decode
  our_pix <- decode_deflate_tile(
    fetched[[i]]$raw_bytes,
    tile_width  = actual_w,
    tile_height = actual_h,
    nbytes      = 2L,
    predictor   = predictor
  )

  ## Apply nodata → NA to match GDAL's convention
  if (!is.na(nodata)) {
    our_pix[our_pix == nodata] <- NA_integer_
  }

  ## GDAL's read
  gdal_pix <- read_block_via_gdal(cog_url, tc, tr)

  results <- rbind(results, data.frame(
    tile_col = tc, tile_row = tr,
    match = identical(as.numeric(our_pix), as.numeric(gdal_pix))
  ))
}

results
cat(sprintf("\n%d / %d tiles match\n", sum(results$match), nrow(results)))
```

## Step 5: Visual overview

The tile grid overlaid on the Sentinel-2 image (overview level 3 for speed),
with match results labelled.

```{r}
#| label: grout-vis
#| fig-width: 8
#| fig-height: 8

r <- terra::rast(sprintf("vrt://%s?ovr=3", cog_url))
terra::plot(log(r))
plot(gtile, add = TRUE)

## Join results to grout's tile index (grout is 1-based, results are 0-based)
gindex2 <- dplyr::left_join(
  gindex,
  results |> dplyr::mutate(tile_col = tile_col + 1L, tile_row = tile_row + 1L),
  c("tile_col", "tile_row")
) |>
  dplyr::filter(!is.na(match))

## Label at tile centres
text(
  x = (gindex2$xmin + gindex2$xmax) / 2,
  y = (gindex2$ymin + gindex2$ymax) / 2,
  labels = gindex2$match,
  col = ifelse(gindex2$match, "green", "red"),
  cex = 0.8,
  font = 2
)
```

## Step 6: Side-by-side pixel comparison

```{r}
#| label: visual-compare
#| fig-width: 10
#| fig-height: 5

## Pick a tile with interesting content — avoid edges and nodata-heavy tiles
## (9,4) or similar from the image centre works well
tc <- 9L
tr <- 4L

ref <- get_block_byte_ref(ds, band = 1, tile_col = tc, tile_row = tr)
raw_viz <- fetch_tile_bytes(cog_url, ref$byte_offset, ref$byte_length)

actual_w <- min(block_size[1], dim_xy[1] - tc * block_size[1])
actual_h <- min(block_size[2], dim_xy[2] - tr * block_size[2])

gdal_pix <- read_block_via_gdal(cog_url, tc, tr)
our_pix  <- decode_deflate_tile(raw_viz, actual_w, actual_h,
                                 nbytes = 2L, predictor = predictor)
if (!is.na(nodata)) our_pix[our_pix == nodata] <- NA_integer_

par(mfrow = c(1, 2))
ximage::ximage(matrix(gdal_pix, actual_h, byrow = TRUE),
      main = sprintf("GDAL vrt://?block=%d,%d", tc, tr),
      zlim = quantile(gdal_pix, c(0, .99), na.rm = TRUE),
      col = hcl.colors(256), axes = FALSE)
ximage::ximage(matrix(our_pix, actual_h, byrow = TRUE),
      main = sprintf("DuckDB cache decode — (%d,%d)", tc, tr),
      zlim = quantile(our_pix, c(0, .99), na.rm = TRUE),
      col = hcl.colors(256), axes = FALSE)
```

## Step 7: Export the cache

```{r}
#| label: export-cache

## Export catalog (without blobs) to Parquet
dbExecute(con, "
  COPY (
    SELECT source_uri, ifd_index, tile_col, tile_row,
           pixel_x_off, pixel_y_off, pixel_x_size, pixel_y_size,
           geo_xmin, geo_ymax, geo_xmax, geo_ymin,
           byte_offset, byte_length,
           octet_length(raw_bytes) as cached_bytes_size
    FROM tile_catalog
  ) TO 'tile_catalog.parquet' (FORMAT PARQUET)
")
```

## Cleanup

```{r}
#| label: cleanup
ds$close()
dbDisconnect(con, shutdown = TRUE)
```

## What this proves

1. **COG structure is transparent** — the tile grid is a queryable catalog, not
   opaque internal state locked inside GDAL's block cache
2. **Bytes are portable** — compressed tile data moves from S3 → DuckDB →
   decoded pixels with no GDAL in the decode path
3. **Cache is inspectable** — SQL queries answer "what's cached?", "how much
   data?", "which tiles cover this geographic extent?"
4. **Cache is shareable** — a `.duckdb` or `.parquet` file replaces "you need
   access to this S3 bucket"
5. **GDAL agreement** — independent byte-level access produces pixel-identical
   results to GDAL's internal pipeline

## Key discoveries along the way

- **`BLOCK_OFFSET_{x}_{y}` in the `"TIFF"` metadata domain**: GDAL's GTiff
  driver exposes per-tile byte offsets as band-level metadata, accessible via
  `gdalraster::$getMetadataItem()`. Two lines of R give you what would
  otherwise require parsing TIFF IFDs.
- **`vrt://?block=i,j`** (GDAL 3.13): Reads a single internal block as a
  subwindow dataset. Clean oracle for comparison without manual window
  arithmetic.
- **`as_blob()` not `blob()`**: The blob package's constructor rejects lists
  of raw vectors; `as_blob()` accepts them.
- **`octet_length()` not `length()`**: DuckDB's SQL function for BLOB sizes.
- **`memDecompress(..., type = "gzip")`**: TIFF DEFLATE uses zlib-wrapped
  deflate, which R handles as gzip. `type = "unknown"` does not work.
- **Predictor=2 on UInt16**: Sample-level horizontal differencing, not
  byte-plane interleaving. Recovery is `cumsum()` per row, mod 65536.
- **Edge tiles**: Bottom and right edge tiles are smaller than the block size
  (e.g. 740 instead of 1024 for a 10980-pixel image with 1024-pixel blocks).
  The decode must use actual tile dimensions.

## Next steps

- [ ] Fetch all 121 tiles (not just 20) and verify 121/121
- [ ] Catalog overview IFDs via `vrt://{url}?overview_level=N`
- [ ] Compute which web map tiles (z/x/y in EPSG:3857) need which source tiles
- [ ] Store a `warp_config` table: source CRS, target CRS, resampling, tile size
- [ ] Transform tile corners through PROJ to validate source→dest mapping
- [ ] Compare with what a Rust `async-tiff` reader fetches for the same tiles
